Exploratory Analysis on CDC Teen Vaccination Surveys

### 2024-05-03 Reviews/project-review-by-mutsa.qmd
Updated manuscript

---
title: Project Review Template 
author: Arlyn Santiago
date: date-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

@@

Title of project: HPV vaccination completion rates depending on several socioeconomic factors

Name of project author(s):Rachel Robertson and Kelly Cao

Name of project reviewer:Arlyn Santiago

@@

I believe both authors do a job of explaining the purpose of this project and providing a solid background on HPV.

@@

### Feedback and Comments
I think the question is clear that their aim is to identify the socioeconomic and geographic factors associated with HPV vaccination completion for teenagers aged 13 to 17 in the U.S.

@@

* question/hypotheses fully explained

@@

I think the data is well explained, however, I think it would be best to include all variable names and meanings into a codebook excel sheet, and place this excel sheet into the data folder. Or, is this the README file located in raw-data folder?

@@

Maybe I wasn't looking in the right file, but in your EDA-code qmd there wasn't much explanation on how the data was cleaned. Most of your cleaning was done in the cleaning1.R file so it would be best to direct your reader to this file. The following is mentioned, "The following data set has been cleaned.Any observations with NAs has been removed." Maybe it would be best to add comments to your code such as using drop_na() to remove NAs. I believe your plots are helpful.

@@

* some weaknesses in wrangling and exploratory component

@@

The analysis was done properly done. However, it was a bit difficult to understand since the analysis was done over seperate files. It would be best to include all the analysis on one file. It's possible I missed it but you could also add Cross Validation to your LASSO to help the optimal lambda value and ensure the model generalizes well.

@@

The results are very well presented, however, I would rename some of the variables such as INCPOV1 your plots to the variable meaning (poverty status). For Figure it would be best to rotate your x-axis labels. This website does a good job explaining how (https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2)

@@

I was not able to find any discussion/conclusion. I would try to interpret the findings of your random forest model as they gave you guys the best findings compared to the other models.

### Summary assessment
* major parts of discussion missing or wrong

@@

_Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project._

@@

I think the project is labeled in a very self-explanatory way, however, it would be best to place your data cleaning in the EDA file or mention the data cleaning within the EDA file. Also, the analysis should be placed within one file for ease of access.

@@

* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

@@

### Feedback and Comments
I think enough information was provided to understand each step of the analysis.

@@

I was able to fully run everything.

@@

* fully reproducible without issues

@@

If both of guys are able to fully address the hypothesis and explain your findings, then you should have an through overall study.

@@

* decent level of thoroughness

@@

You guys have done a great job! I would try and add more information on vaccine bias, especially within marginalized groups and how the findings of this study might combat that in your Discussion.

>>>>

---
title: Project Review Part 5 
author: Mutsa Nyamuranga
date: 04-25-2024
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

@@

Title of project: Teenage HPV Vaccination Coverage and Socioeconomic Factors

Name of project author(s): Kelly Cao, Rachel Robertson

Name of project reviewer: Mutsa Nyamuranga

# Instructions

@@

The background was very thorough with multiple citations and necessary cotectual evidence.

@@

The research question was clearly stated and explained.

@@

* question/hypotheses fully clear

@@

Data source was clearly described. There was a codebook provided.

@@

Data was well processed. Detailed comments/notes were provided for all procedures.
There was no supplementary material.

@@

* essentially no weaknesses in wrangling and exploratory component

@@

The analysis methods were appropriate for the question to be answered. Cross-validation was used for performance evaluation and there were multiple models produced and tested to gather the appropriate metrics.

@@

The results were presented in a well-structured manner.

@@

There was no discussion or conclusion posted

* major parts of discussion missing or wrong

@@

There were some edits that I made to allow code to flow, however, overall the project is well done and very comprehensive.

@@

The project is generally well-structured. All files are in well labeled folders. It is easy to figure out the use for each file/folder.

@@

* well structured

@@

The R scripts are well-documented with sufficient description. Each decision of analysis was justified and explained. The Rmd files(README) are well utilized.

@@

I was able reproduce the results. Instructions were clear.

@@

* small parts not reproducible or required manual intervention

@@

The research questions have been fully addressed. All sections are relatively complete and thorough.

@@

* strong level of thorougness

@@

Overall, this study is in good shape. The authors may want to consider touching up code within some of the R scripts, however they do not take away from how well the project was done.

### 2024-05-03 code/analysis-code/analysis1.qmd
Updated manuscript

```
The residual plot shows clear collinearity that is not represented by the current model. To find the predictors with multicollinearity, we will use a correlation matrix. This is because the VIF analysis will not run when multicollinearity is too high.
```{r}
factor_data <- model.matrix(~ STATE + FACILITY + RACEETHK + AGE + EDUC1 + INS_BREAK_I + INCPOV1 + SEX + LANGUAGE + MOBIL_1 + INS_STAT2_I, data = mydata2)

>>>>

# Save the plot as an image
file_path = here("results","figures", "residual_plot.png")
ggsave(filename = file_path, plot = residual_plot_logit)
```
The residual plot shows clear collinearity that is not represented by the current model. To find the predictors with multicollinearity, we will use a correlation matrix. This is because the VIF analysis will not run when multicollinearity is too high.
```{r}
factor_data <- model.matrix(~ STATE + FACILITY + RACEETHK + AGE + EDUC1 + INS_BREAK_I + INCPOV1 + SEX + LANGUAGE + MOBIL_1 + INS_STAT2_I, data = mydata2)

### 2024-05-03 code/analysis-code/analysis2.5.qmd
Updated manuscript

print(rf_vars)
```
The most important variable, according to this model, is state, followed by income poverty ratio, then facility, then facility. Despite the importance above, it should be noted that the multicollinearity may affect the importance of the variables selected. This means that a model that regularizes the parameters could be ideal for finding the most important predictors.
For this, we will try using the LASSO model.

>>>>

# Modify the plot to set the color of the bars
rf_vars_plot <- rf_vars+
  aes(fill = "skyblue")+
  labs(title = "Top Predictors")

# Original variable names and their translations
variable_translations <- data.frame(
  Original_Variable = c("MOBIL_1", "SEX", "RENT_OWN", "INS_STAT2_I", "EDUC1", 
                         "RACEETHK", "AGE", "FACILITY", "INCQ298A", "STATE"),
  Translated_Variable = c("Geographic Mobility", "Sex", "Rental Status", 
                          "Insurance Status", "Mother's Education Status", "Ethnicity", "Age", 
                          "Facility", "Income", "State")
)

# Add translations to the plot
rf_vars_plot_with_translations <- rf_vars_plot +
  scale_fill_manual(values = c("skyblue")) +  # Set fill color manually
  scale_x_discrete(labels = variable_translations$Translated_Variable) +  # Translate x-axis labels
  labs(title = "Top Predictors",
                x = "Variable")  # Add title and update x-axis label

# Display the plot with translated variable names
print(rf_vars_plot_with_translations)

# Save the plot as a PNG file
file_path = here("results", "figures", "variable_importance_plot.png")
ggsave(filename = file_path, plot=rf_vars_plot_with_translations) 
```
The most important variable, according to this model, is state, followed by income poverty ratio, then facility, then facility. Despite the importance above, it should be noted that the multicollinearity may affect the importance of the variables selected. This means that a model that regularizes the parameters could be ideal for finding the most important predictors.
For this, we will try using the LASSO model.

### 2024-05-03 code/analysis-code/analysis2.qmd
Updated manuscript

#### Loading library and data set
``` {R, warnings=FALSE, messages=FALSE}
# Load required libraries
library(MASS) # for robust regression
library(caret) # for cross-validation
library(h2o) #For dimension reduction
library(dplyr) 
library(rsample)

@@

@@

>>>>

#### Loading library and data set
``` {R, warnings=FALSE, messages=FALSE}
# Load required libraries
library(MASS) # for robust regression
library(caret) # for cross-validation
library(h2o) #For dimension reduction
library(dplyr) 
library(rsample)
library(knitr)
library(kableExtra)
library(webshot)
library(here)
library(gt)
library(gtsummary)
```

@@

```{r}
# Extract the components from the PCA model
standard_deviation <- pca_model@model$model_summary$pc1[1]
proportion_variance <- pca_model@model$model_summary$pc1[2]
cumulative_proportion <- pca_model@model$model_summary$pc1[3]

# Create a data frame with the PC information
scree_data <- data.frame(
  Component = "PC1",
  Standard.deviation = standard_deviation,
  Variance = standard_deviation^2,
  Proportion.Variance = proportion_variance,
  Cumulative.Proportion = cumulative_proportion
)

# Print the summary table
print(scree_data)

# Create a table using kableExtra
pca_table <- 
  scree_data %>%
  gt() %>%
  tab_header(
    title = "PCA Summary Table"
  )

# Save the table to a file
gtsave(pca_table, here("results","tables", "pca_summary.png"))
```

@@

# Function to extract metrics from a model
extract_metrics <- function(model, dataset) {
  metrics <- model@model$training_metrics@metrics
  mse <- metrics$MSE
  r2 <- metrics$r2
  logloss <- metrics$logloss
  mpce <- metrics$mean_per_class_error
  auc <- metrics$AUC
  gini <- metrics$Gini
  nulldev <- metrics$null_deviance
  resdev <- metrics$residual_deviance
  aic <- metrics$AIC

  # Create a data frame for metrics
  metrics_df <- data.frame(
    Metric = c("MSE", "R^2", "Log Loss", "Mean Per-Class Error", "AUC", "Gini", "Null Deviance", "Residual Deviance", "AIC"),
    Value = c(mse, r2, logloss, mpce, auc, gini, nulldev, resdev, aic),
    Dataset = dataset
  )

  return(metrics_df)
}

# Extract metrics from the LASSO model (training data)
train_metrics_df <- extract_metrics(lasso_model, "Training")

# Extract metrics from the CV metrics (test data)
test_metrics_df <- extract_metrics(cv_metrics, "Test")

# Combine the training and test metrics data frames
combined_metrics_df <- rbind(train_metrics_df, test_metrics_df)

# Create a gt table
gt_table <- combined_metrics_df %>%
  gt(
    groupname_col = "Dataset", 
    row_group_as_column = TRUE
  ) %>%
  tab_header(
    title = md("Model Evaluation Metrics"),
    subtitle = md("Comparison of Metrics between Test and Training Datasets")
  )

# Save the gt table
gtsave(gt_table, here("results", "tables", "LASSO_metrics.png"))

### 2024-05-03 code/analysis-code/analysis3.qmd
Updated manuscript

---
title: "Data Analysis and Modeling of Teen Vaccination Surveys"
date: '`r format(Sys.Date())`'
Author: Kelly Cao and Rachel Robertson
output: html_document
editor: 
  markdown: 
    wrap: sentence
---
# LASSO using Tidymodels
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```
We will start by opening the libraries that we will need for creating a LASSO regression model for the data
```{r}
library(tidymodels) # use tidymodels framework
library(ggplot2) # producing visual displays of data
library(dplyr) # manipulating and cleaning data
library(here) # making relative pathways
library(glmnet) # for LASSO regression engine
library(doParallel) # for parallel processing
library(rsample) # for cross validation
library(yardstick) # for metrics
library(mlbench)
library(vip)
```

@@

print(lasso_vars)
```
It seems that the most important factors for HPV vaccine completion are the state, facility and housing status. The states that have the highest predictive power of HPV vaccine completion are Rhode Island, Vermont, Mississippi, and North Dakota.
```{r}
# Extract and tidy model parameters
lasso_tune_coef <- lasso_tune_fit %>%
  extract_fit_parsnip() %>%
  tidy()

>>>>

---
title: "Data Analysis and Modeling of Teen Vaccination Surveys"
date: '`r format(Sys.Date())`'
Author: Kelly Cao and Rachel Robertson
output: html_document
editor: 
  markdown: 
    wrap: sentence
---
# LASSO using Tidymodels
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```
We will start by opening the libraries that we will need for creating a LASSO regression model for the data
```{r}
library(tidymodels) # use tidymodels framework
library(ggplot2) # producing visual displays of data
library(dplyr) # manipulating and cleaning data
library(here) # making relative pathways
library(glmnet) # for LASSO regression engine
library(doParallel) # for parallel processing
library(rsample) # for cross validation
library(yardstick) # for metrics
library(mlbench)
library(vip)
library(gt)
```

@@

# Modify the plot to set the color of the bars
lasso_vars_plot <- lasso_vars+
  aes(fill = "skyblue")+
  labs(title = "Top Predictors")

# Create a data frame with original and translated variable names
variable_translations <- data.frame(
  Original_Variable = rev(c("STATE_RHODE.ISLAND", "FACILITY", 
                        "STATE_VERMONT", "STATE_MISSISSIPPI", "STATE_NORTH.DAKOTA", "STATE_IOWA", 
                        "STATE_HAWAII", "RENT_OWN_DON.T.KNOW", "STATE_NEBRASKA", 
                        "INS_BREAK_I_CURRENTLY.UNINSURED.AND.NEVER.INSURED.SINCE.AGE.11")),
  Translated_Variable = rev(c("Rhode Island", "All Std School Teen Clinics or Other Facilities", 
                          "Vermont", "Mississippi", "North Dakota", "Iowa", "Hawaii", 
                          "Rent/Own Status (Don't Know)", "Nebraska", 
                          "Uninsured"))
)

# Add translations to the plot
lasso_vars_plot_with_translations <- lasso_vars_plot +
  scale_fill_manual(values = c("skyblue")) +  # Set fill color manually
  scale_x_discrete(labels = variable_translations$Translated_Variable) +  # Translate x-axis labels
  labs(title = "Top Predictors",
                x = "Variable")+  # Add title and update x-axis label
  scale_y_continuous(labels = function(x) strwrap(x, width = 10))

# Display the plot with translated variable names
print(lasso_vars_plot_with_translations)

# Save the plot as a PNG file
path = here("results", "figures", "top_lasso_predictors.png")
ggsave(filename = path, plot=lasso_vars_plot_with_translations) 
```
It seems that the most important factors for HPV vaccine completion are the state, facility and housing status. The states that have the highest predictive power of HPV vaccine completion are Rhode Island, Vermont, Mississippi, and North Dakota.
```{r}
# Extract and tidy model parameters
lasso_tune_coef <- lasso_tune_fit %>%
  extract_fit_parsnip() %>%
  tidy()

### 2024-05-03 code/eda-code/EDA-code.qmd
Updated manuscript

```{r}
p1 <- mydata %>% ggplot(aes(x=INCPORAR_I)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures", "income.png")
ggsave(filename = figure_file, plot=p1) 
```

@@

#plot with ggplot
p5 <- ggplot(family_income_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Family Income Distribution Percentage",
       x = "Income",
       y = "Percentage")

@@

#plot with ggplot
p13 <- ggplot(facility_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Facility Distribution Percentage",
       x = "Facility",
       y = "Percentage") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

@@

```{r}
#Stratifying the HPV vaccine completion by facility
##Make a stacked bar plot to compare which facilities reported UTD patients
p18 <- ggplot(mydata, aes(x = P_UTDHPV, fill = FACILITY)) +
  geom_bar(stat = "count") +  # Use geom_bar() with the count of each facility
  labs(x = "Vaccination Status", y = "Count", title = "Facility Vaccination Completion") +
  theme_minimal()
  print(p18)

@@

p17 <- mydata %>% 
  ggplot(aes(x=INS_STAT2_I, y=INCPORAR_I, color = P_UTDHPV)) +  
  geom_boxplot() +
  labs(x = "Insurance status", y = "Income-poverty ratio", title = "Income-poverty ratio stratified by insurance and vaccination") +
  scale_x_discrete(labels = c("PRIVATE INSURANCE ONLY" = "Private", "ANY MEDICAID" = "Medicaid", "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB. WITH PRIVATE INSURANCE)" = "Other", "UNINSURED" = "Uninsured")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))  # Rotate x-axis text by 90 degrees
plot(p17)
figure_file = here("results","figures", "insurance-income-stratified.png")
ggsave(filename = figure_file, plot=p17)

@@

### Fit 4
```{r}
#Stratifying the HPV vaccine completion by both Race-Ethnicity and (choose either insurance or income-poverty status)
p19 <- mydata %>% 
  ggplot(aes(x=INCPORAR_I, y=RACEETHK, color = P_UTDHPV)) +  geom_boxplot()
plot(p19)
figure_file = here("results","figures", "raceeth-income-stratified.png")
ggsave(filename = figure_file, plot=p19)

>>>>

```{r}
p1 <- mydata %>% 
  ggplot(aes(x = INCPORAR_I)) +  
  geom_histogram(binwidth = 1, color = "black", fill = "skyblue") +  # Adjust binwidth as needed
  xlab("Income to Poverty Ratio") +                  
  ylab("Count") +                                    
  ggtitle("Income-to-Poverty Ratio Variable Distribution") +
  theme_minimal() +  # Apply a minimal theme for cleaner appearance
  theme(
    plot.title = element_text(size = 14, face = "bold"),  # Customize title font size and style
    axis.title.x = element_text(size = 12),              # Customize x-axis label font size
    axis.title.y = element_text(size = 12),              # Customize y-axis label font size
    axis.text = element_text(size = 10),                 # Customize axis text font size
    legend.position = "none",                            # Remove legend if not needed
    panel.grid.major = element_line(color = "lightgray"), # Add light gray grid lines
    panel.grid.minor = element_blank()                   # Remove minor grid lines
  ) +
  labs(caption = "(Originally INCPORAR_I)")

p1
figure_file = here("results", "figures", "income.png")
ggsave(filename = figure_file, plot=p1) 
```

@@

#plot with ggplot
p5 <- ggplot(family_income_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +  # black borders
  labs(title = "Household Income Variable Distribution",
       x = "Income",
       y = "Percentage (%)") +
  theme_minimal() +  # Minimalistic theme
  labs(caption = "(Originally INCQ298A)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),  # Rotate x-axis labels
        axis.line = element_line(color = "black"),           # Black axes
        panel.grid.major = element_blank(),                  # Remove major gridlines
        panel.grid.minor = element_blank(),                  # Remove minor gridlines
        panel.border = element_blank())                      # Remove panel borders

@@

#plot with ggplot
p13 <- ggplot(facility_df, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Facility Distribution Percentage",
       x = "Facility",
       y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5))

@@

```{r}
#Stratifying the HPV vaccine completion by facility
##Make a stacked bar plot to compare which facilities reported UTD patients
p18 <- ggplot(mydata, aes(x = P_UTDHPV, fill = FACILITY)) +
  geom_bar(stat = "count") +  
  labs(x = "Vaccination Status", y = "Count", caption = "UTD = Up-to-Date", title = "Facility Vaccination Completion") +
  theme_minimal()
  print(p18)

@@

p17 <- mydata %>% 
  ggplot(aes(x = INS_STAT2_I, y = INCPORAR_I, color = P_UTDHPV)) +  
  geom_boxplot() +
  labs(x = "Insurance status",
       y = "Income-to-Poverty ratio",
       title = "Income-to-Poverty Ratio Stratified by Insurance and Vaccination Status", 
       color = "Vaccination Status", 
       caption = "UTD refers to Up-to-Date vaccination status") +  
  scale_x_discrete(
    labels = c("PRIVATE INSURANCE ONLY" = "Private",
               "ANY MEDICAID" = "Medicaid",
               "OTHER INSURANCE (CHIP, IHS, MILITARY, OR OTHER, ALONE OR IN COMB. WITH PRIVATE INSURANCE)" = "Other",
               "UNINSURED" = "Uninsured")) +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5))  # Rotate x-axis text by 90 degrees

print(p17)

figure_file = here("results","figures", "insurance-income-stratified.png")
ggsave(filename = figure_file, plot=p17)

@@

### Fit 4
```{r}
#Stratifying the HPV vaccine completion by both Race-Ethnicity and (choose either insurance or income-poverty status)
p19 <- mydata %>% 
  ggplot(aes(x=INCPORAR_I, y=RACEETHK, color = P_UTDHPV)) +
  geom_boxplot()+
  labs(title = "HPV Vaccination Status",
       x = "Income-to-Poverty Ratio",
       y = "Race/Ethnicity", 
       caption = "A ratio equal to 1, suggests income and poverty levels are the same.", 
       color = "Vaccination Status")
plot(p19)
figure_file = here("results","figures", "raceeth-income-stratified.png")
ggsave(filename = figure_file, plot=p19)
